Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-02-09 paper/main.tex
rewrite of intro/and consideration of pulling up definitions

\IEEEPARstart{V}isualizations are expected to be a faithful translation of data, meaning that inherent structure in the data should be present in the visualization; therefore the components of visualization software libraries that translate data to graphics are also expected to be structure preserving. This inherent structure can be quite complex, such as when the data is high dimensional, multivariate, deeply nested, or encoded in a distributed manner. We propose that we can rigorously specify what structure the library components are expected to preserve using category theory. We also propose that these specifications can generalize to most structure types by extending previous work on encoding  inherent structure in a consistent manner using mathematical structures from algebraic topology. These algebraic structures can be translated into analogous programmatic types for encoding structure, while the use of category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e \in G$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.

@@

 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}
For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings.

The notion that visual encodings should be \textit{homomorphic} was proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} in his specification of \textit{A Presentation Tool} and the idea that visual transforms are \textit{equivariant} underlies Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. Kindlemann and Scheidegger propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

>>>>

\note{The mapping between data and visualization is expected to be bidirectional -> that's what we really mean by equivariance and homomorphism  }
\IEEEPARstart{V}isualizations are expected to be structure preserving, meaning that when data is transformed into a graphical representation enough of the structure of the data is present in the graphic that the data can be distilled solely from the graphical representation. Changes to structure through changing the data or the graphic in the $data\leftrightarrow graphic$ mapping are expected to be reflected in each inputs corresponding output \cite{kindlmannAlgebraicProcessVisualization2014}.Because visualizations are assumed to preserve structure, visual algorithms implemented in visualization producing tools, such as software libraries, are also expected to preserve structure; frequently these expectations are either codified implicitly, only for some aspects of the data structure, or inconsistently with respect to visualization algorithm type. This means that currently many visualization tools lack an explict, comprehensive, and consistent way to express what conditions must be met to gaurantee that the component produced a structure preserving visualization.

We propose that we can encode the structure of the data and graphics in a uniform generalizable manner by using a mathematical structure from algebraic topology called a fiber bundle, and that we can use category theory to explicitly specify what it means for a visualization tool component to preserve structure. These algebraic structures can be translated into analogous programmatic types for encoding structure, while the use of category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. \note{add in Scales/APT/ GoG/Hullman/ETC}

\subsection{General Structure Preservation}
\note{maybe at top of section?}
Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e \in G$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.

@@

 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}

The notion that visual encodings should be \textit{homomorphic} was proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} in his specification of \textit{A Presentation Tool} and the idea that visual transforms are \textit{equivariant} underlies Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. Kindlemann and Scheidegger propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. A uniform abstract data representation layer is introduced in \autoref{sec:atct:sheaves}, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the the correspondence between data and visual changes is included in the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist}. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.
